---
title: "Computer Assignment 3"
author: "Gabriel Lindqvist, Jiahui Li"
date: "2024-12-05"
output: pdf_document
---

Wermuth (1976,Biometrics)reports data collected from a birth clinic,which includes information on the mother's age,her smoking habits (number of cigarettes per day),gestational age (in days)and the survival status of the child.Data were collected during a given period and neither the total sample size nor any margins were held fixed.The aim of the assignment is to find a log-linear model that can be used to gain an understanding of the association between the variables.The model should fit the data well but not be too complex to interpret.

# Exercise3:1

## Task1
In order to find a 'good'model, several models have to be fitted.Start with the saturated model (which has a perfect fit)and remove interaction terms in
a systematic way,where higher order interactions are removed before lower order interactions.No main effect should be removed,since the interest here is the association between the variables.The goodness-of-fit of the different models should be evaluated with deviance (compared to the saturated model)and AIC,and the table below should be completed with the calculated values.
The variable names to be used are:X=Mother's age,Y=Smoking habits, Z=Gestational age and V Child survival.
### 1.read the data
```{r}
data3<-read.csv("data_ca3.csv")
head(data3)
```
### 2. Fitting a loglinear model
##### 2.1 First we start with the saturated model
```{r}
msat<-glm(n~x*y*z*v, family=poisson(link=log), data=data3)
summary(msat)
```
#### **Saturated Model (`msat`)**
- The saturated model includes all variables (\(x, y, z, v\)) and all interaction terms up to the fourth order.
- **Residual Deviance**: \(2.3315 \times 10^{-14}\) with 0 degrees of freedom, indicating a perfect fit. This is expected for a saturated model, as it uses all available degrees of freedom to capture the variability in the data.
- **AIC**: 123.97. AIC is primarily used for model comparison; for now, this value serves as a baseline.

**Significant terms**:
- Significant main effects: \(y\), \(z\), \(v\).
- Significant interactions: \(x:v\), \(z:v\).

**Non-significant terms**:
- Higher-order interactions (\(x:y:z:v\)) and some lower-order interactions (\(x:y, y:z\), etc.) are not significant (\(p > 0.05\)).

**Conclusion**:
The saturated model perfectly fits the data but is overly complex, containing many non-significant interactions. Simplification is necessary to improve interpretability.

#### 2.2 Then all three-way interactions:
```{r}
m3<-glm(n~(x*y*z+x*y*v+x*z*v+y*z*v), family=poisson(link=log), data=data3)
summary(m3)
```
#### **Model with all three-way interactions (`m3`)**
- This model retains only up to three-way interactions, removing the four-way interaction (\(x:y:z:v\)).
- **Residual Deviance**: 0.35935 with 1 degree of freedom. The residual deviance is very low, indicating the model still fits the data well despite simplification.
- **AIC**: 122.33, which is lower than the saturated model's AIC (123.97). This suggests the three-way interaction model is more parsimonious while still providing a good fit.

**Significant terms**:
- Significant main effects: \(y\), \(z\), \(v\).
- Significant interactions: \(x:v\), \(z:v\).

**Non-significant terms**:
- Three-way interactions (\(x:y:z, x:y:v, x:z:v, y:z:v\)) and some lower-order interactions (\(y:z, x:z\)) remain non-significant.

**Conclusion**:
The three-way interaction model provides a simpler, nearly equivalent fit compared to the saturated model. Further simplification can be considered by removing non-significant terms.

#### 2.3 Then we try the two-way interactions and the simplest model with only main effects
```{r}
#Build the two-way interactions
m2 <- glm(n ~ x*y + x*z+ x*v + y*z + y*v + z*v, family = poisson(link = log), data = data3)
summary(m2)
```

```{r}
# Build the simplest model with only main effects
m1 <- glm(n ~ x + y + z + v, family = poisson(link = log), data = data3)
summary(m1)
```
### 3. Stepwise model selection, using AIC

**Perform Stepwise Selection Based on AIC**:
   - Using the `step()` function, we performed stepwise selection to find a balance between model complexity and fit:
     - **Starting Point**: The most complex model (`saturated`).
     - **Scope**: From the simplest model (`m1`) to the most complex model (`saturated`).
     - **Criteria**: AIC (Akaike Information Criterion), which evaluates the trade-off between goodness-of-fit and model complexity.
     - **Direction**: Both forward and backward, allowing variables to be added or removed iteratively.

#### 3.1 First we try from m1 to saturated model, adding a variable each time
```{r}
mstep_AIC_msatm1_forward<-step(m1, direction="forward", trace=TRUE, scope = list(upper = msat, lower = m1))
summary(mstep_AIC_msatm1_forward)
```
##### **Final Model (`mstep_AIC`)**
1. **Formula**:
   \[
   n \sim x + y + z + v + x:y + x:z + x:v + y:v + z:v
   \]
   - Includes main effects: \(x, y, z, v\).
   - Retains significant two-way interactions: \(x:y, x:z, x:v, y:v, z:v\).
   - Excludes all three-way and higher-order interactions.

2. **Fit and Complexity**:
   - **Residual Deviance**: 1.8221 with 6 degrees of freedom (excellent fit).
   - **AIC**: 113.8 (lowest among tested models).
   - **Significance**: All retained terms are significant or borderline significant, ensuring interpretability.

#### 3.2 Next try from saturated model to m1, removing a variable each time
```{r}
mstep_AIC_msatm1_backward<-step(msat, direction="backward", trace=TRUE, scope = list(upper = msat, lower = m1))
summary(mstep_AIC_msatm1_backward)
```
##### **Final Model (`mstep_AIC`)**
1. **Formula**:
   \[
   n \sim x + y + z + v + x:y + x:z + x:v + y:v + z:v
   \]
   - Includes main effects: \(x, y, z, v\).
   - Retains significant two-way interactions: \(x:y, x:z, x:v, y:v, z:v\).
   - Excludes all three-way and higher-order interactions.

2. **Fit and Complexity**:
   - **Residual Deviance**: 1.8221 with 6 degrees of freedom (excellent fit).
   - **AIC**: 113.8 (lowest among tested models).
   - **P value**: 0.94, which means we could accept this model

#### 4. The final sheet——from m1 to saturated model
| Model                                                                                     | Deviance  | df | AIC    | p_value      |
|-------------------------------------------------------------------------------------------|-----------|----|--------|--------------|
| x * y * z * v (saturated model)                                                          | 0         | 0  | 123.97 | NA           |
| x + y + z + v + x:y + x:z + y:z + x:v + y:v + z:v + x:y:z + x:y:v + x:z:v + y:z:v        | 0.35935   | 1  | 122.33 | 0.55         |
| x + y + z + v + x:y + x:z + y:z + x:v + y:v + z:v + x:y:v + x:z:v + y:z:v                | 0.41133   | 2  | 120.39 | 0.81         |
| x + y + z + v + x:y + x:z + y:z + x:v + y:v + z:v + x:y:v + x:z:v                        | 0.63838   | 3  | 118.61 | 0.89         |
| x + y + z + v + x:y + x:z + x:v + y:v + z:v + x:y:v + x:z:v                              | 0.75076   | 4  | 116.72 | 0.94         |
| x + y + z + v + x:y + x:z + x:v + y:v + z:v + x:z:v                                      | 1.30759   | 5  | 115.28 | 0.93         |
| x + y + z + v + z:v + x:y + x:v + y:v + x:z                                              | 1.8221    | 6  | 113.8  | 0.94         |
| x + y + z + v + z:v + x:y + x:v + y:v                                                   | 4.7486    | 7  | 114.72 | 0.69         |
| x + y + z + v + z:v + x:y + x:v                                                         | 7.7197    | 8  | 115.69 | 0.46         |
| x + y + z + v + z:v + x:y                                                               | 17.845    | 9  | 123.82 | 0.037        |
| x + y + z + v + z:v                           | 35.45     | 10 | 139.43 | 1.046e-4     |
| x + y + z + v                                 | 377.79    | 11 | 479.76 | 0            |

## Task2
Choose from your table a model with few parameters and a good fit.Describe the procedure to compare different models.

### The best model chosen from task1
```{r}
mbest <- glm(n ~ x + y + z + v + x:y + x:z + x:v + y:v + z:v, family = poisson(link = log), data = data3)
summary(mbest)
```
### LR-test between two nested models
```{r}
anova(mbest,msat,test="LRT")
```
From the result, we can see that:

**Fit and Complexity**:
   - **Residual Deviance**: 1.8221 with 6 degrees of freedom (excellent fit).
   - **P value**: 0.94, which means we could accept this model with a good fit.

### **Procedure to Compare Models**

1. **Define Candidate Models**:
   - **Most Complex Model (`saturated`)**: Includes all main effects and interactions:
     \[
     n \sim (x \times y \times z  \times v )
     \]
   - **Simplest Model (`m1`)**: Includes only the main effects:
     \[
     n \sim x + y + z + v
     \]

2. **Perform Stepwise Selection Based on AIC**:
   - Using the `step()` function, we performed stepwise selection to find a balance between model complexity and fit:
     - **Starting Point**: The most complex model (`saturated`).
     - **Scope**: From the simplest model (`m1`) to the most complex model (`saturated`).
     - **Criteria**: AIC (Akaike Information Criterion), which evaluates the trade-off between goodness-of-fit and model complexity.
     - **Direction**: Both forward and backward, allowing variables to be added or removed iteratively.

3. **Evaluate Models Using Residual Deviance and AIC**:
   - For each model, the residual deviance (measuring the lack of fit) and AIC were calculated.
   - Models with lower residual deviance fit the data better, and models with lower AIC balance fit and simplicity.

4. **Compare Models Using Statistical Tests**:
   - **Likelihood Ratio Test (LRT)**:
     - Compared nested models (e.g., `saturated` vs. `m1`) to assess whether removing higher-order terms significantly worsens the fit.
     - Significant \(p\)-values (\(p < 0.05\)) indicate that removing terms leads to a significant loss of fit. \( P = 1 - \text{chi2.cdf(Deviance, Df)} \)

5. **Select the Final Model**:
   - The final model (`mstep_AIC`) was chosen based on its low AIC, good residual deviance, and simplicity. It balances fit and complexity effectively.
In this example, the saturated model is the most complex model that can be considered and m1 is the simplest.

## Task3
Interpret the model you chose. Which associations are significant? Quantify the associations with odds ratios together with confidence intervals.

### Interpret the model we chose
```{r}
summary(mbest)
```
#### **1. Chosen Model**
\[
n \sim x + y + z + v + x:y + x:z + x:v + y:v + z:v
\]

##### Variables:
- \(x\): Mother's age.
- \(y\): Smoking habits.
- \(z\): Gestational age.
- \(v\): Child survival.
- Significant two-way interactions:
  - \(x:y\): Interaction between mother's age and smoking habits.
  - \(x:z\): Interaction between mother's age and gestational age.
  - \(x:v\): Interaction between mother's age and child survival.
  - \(y:v\): Interaction between smoking habits and child survival.
  - \(z:v\): Interaction between gestational age and child survival.

---

#### **2. Significant Associations**

##### **Main Effects**:
1. **\(x\) (Mother's Age)**:
   - Coefficient: \(-0.29199\), \(p = 0.08648\) (marginal significance).
   - Interpretation: Increasing mother's age slightly decreases the likelihood of the outcome.

2. **\(y\) (Smoking Habits)**:
   - Coefficient: \(-1.71313\), \(p < 0.001\).
   - Interpretation: Smoking significantly decreases the likelihood of the outcome.

3. **\(z\) (Gestational Age)**:
   - Coefficient: \(-0.77237\), \(p < 0.001\).
   - Interpretation: Shorter gestational age significantly decreases the likelihood of the outcome.

4. **\(v\) (Child Survival)**:
   - Coefficient: \(1.81422\), \(p < 0.001\).
   - Interpretation: Child survival significantly increases the likelihood of the outcome.

---

##### **Two-Way Interactions**:
1. **\(x:y\)**:
   - Coefficient: \(-0.41132\), \(p < 0.001\).
   - Interpretation: The negative effect of smoking is stronger for older mothers.

2. **\(x:z\)**:
   - Coefficient: \(-0.16557\), \(p = 0.08456\) (marginal significance).
   - Interpretation: The relationship between gestational age and outcome weakens slightly for older mothers.

3. **\(x:v\)**:
   - Coefficient: \(-0.46481\), \(p = 0.00983\).
   - Interpretation: The positive effect of child survival is weaker for older mothers.

4. **\(y:v\)**:
   - Coefficient: \(-0.44375\), \(p = 0.06977\) (marginal significance).
   - Interpretation: The positive effect of child survival is weaker for smokers.

5. **\(z:v\)**:
   - Coefficient: \(3.31135\), \(p < 0.001\).
   - Interpretation: The interaction between gestational age and child survival is highly significant, indicating that survival outcomes improve strongly with longer gestational age.

### Quantify the associations with odds ratios together with confidence intervals.
```{r}
# Calculate Odds Ratios and Confidence Intervals
exp_coef <- exp(coef(mbest)) # Odds Ratios
confint_vals <- confint(mbest) # Confidence Intervals
exp_confint <- exp(confint_vals) # Exponentiate to get OR CIs

# Combine results into a table
results <- data.frame(
  Term = names(exp_coef),
  Odds_Ratio = exp_coef,
  CI_Lower = exp_confint[, 1],
  CI_Upper = exp_confint[, 2]
)

# Display results
print(results)
```
### **Key Findings**
1. **Main Effects**:
   - Smoking (\(y\)), gestational age (\(z\)), and child survival (\(v\)) are strongly associated with the outcome, with significant odds ratios.
   - Mother's age (\(x\)) has a weaker, marginally significant association.

2. **Interactions**:
   - Significant interactions (\(x:y, x:v, z:v\)) suggest complex relationships between variables:
     - Smoking's negative effect increases with maternal age.
     - The positive effect of child survival decreases with maternal age and smoking.
     - Gestational age strongly amplifies the effect of child survival, highlighting its critical role.

3. **Statistical Significance**:
   - Terms with confidence intervals that do not include 1 are statistically significant. Marginal associations (\(x, x:z, y:v\)) should be interpreted cautiously.
